Global search algorithms for one-dimensional multiextremal optimization problems with nonlinear constraints

### 2017-09-14 2017/ParOptBook/Chapter4.tex
parOptBook: update chapter 4

%% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\documentclass[graybox]{svmult}
%
%% choose options for [] as required from the list
%% in the Reference Guide
%
%\usepackage{mathptmx}       % selects Times Roman as basic font
%\usepackage{helvet}         % selects Helvetica as sans-serif font
%\usepackage{courier}        % selects Courier as typewriter font
%\usepackage{type1cm}        % activate if the above 3 fonts are
%                           % not available on your system
%
%\usepackage{makeidx}         % allows index generation
%\usepackage{graphicx}        % standard LaTeX graphics tool
%                             % when including figure files
%\usepackage{multicol}        % used for the two-column index
%\usepackage[bottom]{footmisc}% places footnotes at page bottom
%
%% see the list of further useful packages
%% in the Reference Guide
%
%\makeindex             % used for the subject index
%                       % please use the style svind.ist with
%                       % your makeindex program
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\begin{document}

@@

\section{Problem statement}
In the present chapter, the one-dimensional problems of the constrained optimization will
be considered.

>>>>

%% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\documentclass[graybox]{svmult}
%
%% choose options for [] as required from the list
%% in the Reference Guide
%
%\usepackage{mathptmx}       % selects Times Roman as basic font
%\usepackage{helvet}         % selects Helvetica as sans-serif font
%\usepackage{courier}        % selects Courier as typewriter font
%\usepackage{type1cm}        % activate if the above 3 fonts are
%                           % not available on your system
%
%\usepackage{makeidx}         % allows index generation
%\usepackage{graphicx}        % standard LaTeX graphics tool
%                             % when including figure files
%\usepackage{multicol}        % used for the two-column index
%\usepackage[bottom]{footmisc}% places footnotes at page bottom
%\usepackage{mathtools}
%\mathtoolsset{showonlyrefs}
%
%% see the list of further useful packages
%% in the Reference Guide
%
%\makeindex             % used for the subject index
%                       % please use the style svind.ist with
%                       % your makeindex program
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\begin{document}

@@

\section{Problem statement}
In the present chapter, the one-dimensional problems of the constrained optimization will
be considered. Let $\varphi(x)$ and $g_j(x)\le 0,1\le j\le m$ be the real functions defined in a closed interval $[a,b]$ of the real axis. It is requires to find the values $x^*$ satisfying the condition
\begin{equation}
  \label{eq4:problem}
  \varphi(x^*)=\min\{\varphi(x):x\in [a,b], g_j(x)\le 0,1\le j\le m\}
\end{equation}

As before, the objective function $\varphi$ (hereinafter denoted also as $g_{m+1}$) and the left parts of the constraints $g_j(x)\le 0,1\le j\le m$ re assumed to satisfy \emph{Lipschitz condition} with the corresponding constants $L_j,1\le j\le m+1$, namely:
\begin{equation}
  |g_j(x_1)-g_j(x_2)|\le L_j|x_1-x_2|,1\le j\le m+1,x_1,x_2
\end{equation}

\example{
\label{ex4:problem}
Let us consider a problem of type \eqref{eq4:problem} where $x\in [0.6;2.2],m=2$,
\begin{gather}
  \varphi(x)=\cos(18x-3)\sin(10x-7)+1.5, \\
  g_1(x)=\exp(-x/2)\sin(6x-1.5), \\
  g_2(x)=|x|\sin(2\pi x-0.5).
\end{gather}
he exact solution of the problem is $x^*=2.0795,\varphi(x^*)=0.565$. The graphs of the functions $g_1(x),g_2(x)$,and $\varphi(x)$ are presented in Fig. \ref{fig:4_1}.
\begin{figure}[h]
  \label{fig:4_1}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_1.png}
  \caption{One-dimensional global optimization problem}
\end{figure}
}

\section{Partial computability and index scheme of the account of constraints}
Problem \eqref{eq4:problem} can be considered in the statement when each function $g_j,1\le j\le m+1$ is defined and computable only in the corresponding subdomains $Q_j\subset [a,b]$ where
\begin{equation}
  \label{eq4:q}
  Q_1=[a,b],Q_(j+1)=\{x\in Q_j:g_j(x)\le 0\},1\le j\le m.
\end{equation}

For example, in the optimal design problems some characteristics of the technical systems may appear to be undefined if the conditions of the system functioning represented by a part of the constraints of problem \eqref{eq4:problem} are not fulfilled.

Taking into account conditions \eqref{eq4:q}, initial problem \eqref{eq4:problem} can be represented in the form
\begin{equation}
  \varphi(x^*)=\min\{g_{m+1}(x):x\in Q_{m+1}\}.
\end{equation}

For the purposes of further treatment, let us introduce a classification of the points $x$ in the search domain $[a,b]$ using an index $\nu=\nu(x)$ where $\nu-1$ is the number of constraints, which are fulfilled in this point. Formally, the index $\nu$ is defined by the conditions
\begin{equation}
  g_j(x)\le 0,1\le j \le \nu-1,g_\nu(x)>0,
\end{equation}
(the last inequality is inessential  if $\nu=m+1$) and it satisfies the inequalities
\begin{displaymath}
  1\le\nu=\nu(x)\le m+1.
\end{displaymath}
The problem of example \ref{ex4:problem} assuming a partial computability of the functions is presented in Fig. \ref{fig:4_2}. The arcs of the restrictions $g_1(x),\:g_2(x)$, and of the objective function $\varphi(x)$ defined in the corresponding subdomains $Q_j$ from \eqref{eq4:q} are shown. The point $x^1$ with the index $\nu(x^1)=1$, the point $x^2$ with the index $\nu(x^2)=2$, and the point $x^3$ with the index $\nu(x^3)=3$ are shown also.

\begin{figure}[h]
  \label{fig:4_2}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_2.png}
  \caption{A problem with partially computable functions}
\end{figure}

This classification generates the function
\begin{equation}
  \label{eq4:fx}
  f(x)=g_\nu(x),\nu=\nu(x),
\end{equation}
defined and computable everywhere in $[a,b]$. Its value at a point $x$ is either the value of the left part of the constraint violated at this point (the case $\nu\le m$)  or the value of the objective function (the case $\nu=m+1$). Therefore, the determination of the value of $f(x),\:x\in[a,b]$  is reduced to a successive computations of the values of $g_j(x), 1\le j\le \nu=\nu(x)$, i.e., the next value $g_{j+1}(x)$ is calculated in the case when $g_j(x)\le 0$ only. The computational process finishes either as a result of the fulfillment of the inequality $g_j(x)>0$ or as a result of achievement of the value $\nu(x)=m+1$.

The described procedure called a \emph{trial} at the point $x$ forms the index $\nu$ of this point automatically. The pair of values
\begin{equation}
  z=f(x)=g_\nu(x),\nu=\nu(x)
\end{equation}
generated by the trial at the point $x\in[a,b]$ is called the \emph{trial result}.

The graph of the function $f(x)$ from \eqref{eq4:fx}, which consists of the arcs of the constraints $g_1(x),\:g_2(x)$, and of the objective function $\varphi(x)$ taken from Example \ref{ex4:problem} is shown in Fig. \ref{fig:4_3}.

\begin{figure}[h]
  \label{fig:4_3}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_3.png}
  \caption{An «index» function}
\end{figure}
